Boosted Regression Trees for Data-Driven Auto-Tuning

### 2011-12-11 paper/draft1.tex
paper

\begin{figure}[!ht]
\centerline{
\begin{algorithm}{$\Algo{\color{darkred}{gpu\_fbcorr\_empirical}}\big(shapes, strides \big)$}
\Aitem $a \setto TaskFeatures(shapes, strides)$
\Aitem $c \setto PlatformFeatures()$
\Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} MeasureTime(a, b, c)$
\algoremark{\color{darkred}{slow}}
\Aitem \Return $b^*$
\end{algorithm}
}
\begin{algorithm}{$\Algo{\color{darkred}{gpu\_fbcorr\_predictive}}\big(shapes, strides \big)$}
\Aitem $a \setto TaskFeatures(shapes, strides)$
\Aitem $c \setto PlatformFeatures()$
\Aitem $f \setto LoadTimingModel()$
\Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} f(a, b, c)$
\algoremark{\color{darkred}{fast}}
\Aitem \Return $b^*$
\end{algorithm}
\caption{Pseudo-code template for an auto-tuning implementation of
$\texttt{gpu\_fbcorr}$. Empirical auto-tuning (above) is inevitably slow
because dynamically-generated code must be compiled and run on a number of actual-size inputs.
Predictive auto-tuning (below) can be orders of magnitude faster. We show that it can
also be accurate.
}
\label{fig:general_autotune}
\end{figure}

This work shows that auto-tuning can be accelerated by orders of
magnitude by using a regression model built offline as a surrogate for actual
computations on the real hardware.
The general form of an auto-tuning based library routine is illustrated in Figure~\ref{fig:general_autotune} (top).
We will use filterbank correlation (Section~\ref{sec:fbcorr}) as our running example, but the approach is both general and powerful.
An auto-tuning based routine must operate on three sets of variables:
\begin{description}
\item[$\mathcal{A}$:] task description (argument shapes, physical layout)
\item[$\mathcal{B}$:] implementation description (auto-tuning parameters)
\item[$\mathcal{C}$:] platform description (compute capability, micro-benchmarks)
\end{description}
The hypothetical auto-tuning routine described at the top of Figure~\ref{fig:general_autotune} 
might take many minutes or hours to perform the argmin at step 3 (during which time it computes the desired result many times!)
so it would be all but unusable as a normal library routine.
However, the form of the auto-tuning routine suggests the potential for enormous acceleration:
if only there were a fast (even approximate) surrogate for the costly $MeasureTime(\cdot)$ function,
then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~{\ref:fig:general_autotune}, bottom).

Our contribution of the present work is to show that boosted regression tree models,
a powerful machine learning technique for function approximation, can meet the
requirements of this application.
We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning. XXX numbers:
how good, how fast can prediction be done how much training data is necessary?

\section{Filterbank Correlation}
\label{sec:fbcorr}

Filterbank correlation is a simple spatial image filtering operation that is
an important subroutine in many image processing applications. It has a relatively
high arithmetic intensity which makes it a natural fit for GPU platforms.
One might expect then, that it would be easy to implement a library providing
this operation -- this hypothetical library could have quite a simple interface
following the spirit of FFTW or a BLAS routine:
\begin{verbatim}
    gpu_fbcorr(
        img_shape, filters_shape,
        img, filters, output)
\end{verbatim}
However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
related stencil operations XXX, it is challenging to provide an implementation
or even an implementation strategy that provides satisfactory performance
across the range of inputs (shapes, physical layouts) that occur even in
typical usage.

@@

\subsection{Filter-bank Correlation}

Following \citet{pinto+cox:2011gcg}, the experiments in this paper will focus on filterbank
correlation.  Mathematically, we define filterbank correlation in terms of an
image $x$ and a filterbank $f$.
The image $x$ has $R$ rows, $C$ columns, and $D$ channels (e.g. color
channels) that we call its {\em depth}. We index $x$ like $\mathbf{x}[i,j,d]$
where $0 \leq i < R$, $0 \leq j < C$, and $0 \leq d < D$.
The filterbank $f$ has $F$ filters that are like little images: each has a
height $H$, a width $W$, and $D$ channels.
We will restrict ourselves to what are called {\em valid} correlations, in
which the image is larger in both rows and columns than the filters.
The result of filterbank correlation of $x$ with $f$ is an image-like array
$z$ with $R-H+1$ rows, $C-W+1$ columns, and depth $F$, whose elements are
defined according to Equation~\ref{eq:z}:

@@

Filterbank correlation is used extensively in image and video processing,
where it is often the computational bottleneck.
GPUs are well-suited to computing filterbank correlations, although the
optimal strategy for doing so often depends on the shape of the problem, i.e.
the values of $R$, $C$, $D$, $F$, and $W$.

In terms of floating point operations, a filterbank correlation requires the
inner sums to be computed for each output pixel, yielding the quantity in
Eq.~\ref{eq:flops}:
\begin{equation}
\mathrm{FLOPS} = 2  K  H  W  D  (R - H + 1)( C- W + 1)
\label{eq:flops}
\end{equation}
The multiplicative factor of 2 arises because we must multiply an element of
$x$ with an element of $f$ and then add the result to an element of $z$.

In terms of bandwidth, if we were to assume (optimistically) that the entire
filterbank fit into the GPU's so-called constant memory, then the bandwidth
required for a filterbank correlation would be the size of the filterbank,
plus the size of the input image, plus the size of the output image. This sum
is expressed in terms of configuration parameters in Eq.~\ref{eq:bytes}:
\begin{align}
\mathrm{Bytes} = 4&RCD \nonumber \\
& + 4KHWD \nonumber \\
& + 4(R-H+1)(C-W+1)K
\label{eq:bytes}
\end{align}
We will see the bandwidth requirements of a more realistic analysis in
Section~XXX.

In terms of artithmetic density, it's approximately
\begin{align}
\mathrm{intensity} & \approx \frac {KDHW} {2(D+K)},
\end{align}
because each $K$ output writes corresponds to approximately $D$ input reads
and an inner product between $KDHW$ elements.

@@

\section{Configuration Spaces}

Recall from the introduction (Eq.~\ref{eq:condopt}) that we are addressing a
conditional optimization problem, in which we condition on a problem
configuration ($P$) and a system specification ($S$), and search for the best
implementation $I$ for addressing problem $P$ on system $S$.
This section describes the configuration spaces we used in our tests,
which are described later in Section XXX.

\subsection{Problem Description Space}

Mathematically, filterbank correlation is an operation on an $R \times C \times
D$ image and a $F \times H \times W \times D$ filterbank.
We randomly sampled problem specifications from the following space:
\begin{align*}
R = C & \in \{ 256, 512, 1024, 2048, 4096 \} \\
H = W & \in \{ 3, 5, 7, 9, 11 \} \\
D &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \} \\
F &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \}
\end{align*}
The product space
includes 1600 elements, but we restricted our experiments to problems that
represented between 1 and 50 gigaflops. Our problem space included 602
configurations. A library implementation that must support all image and
filter sizes must handle on the order of $2^{50}$ forseable
problem configurations (not even including variations due to strided memory
layouts).

\subsection{Implementation Description Space}

@@

@@

>>>>

\begin{figure}[!ht]
\centerline{
\begin{algorithm}{$\Algo{\color{darkred}{autotune\_empirical}}\big(shapes, strides \big)$}
\Aitem $a \setto TaskFeatures(shapes, strides)$
\Aitem $c \setto PlatformFeatures()$
\Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} MeasureTime(a, b, c)$
\algoremark{\color{darkred}{slow}}
\Aitem \Return $b^*$
\end{algorithm}
}
\begin{algorithm}{$\Algo{\color{darkred}{autotune\_predictive}}\big(shapes, strides \big)$}
\Aitem $a \setto TaskFeatures(shapes, strides)$
\Aitem $c \setto PlatformFeatures()$
\Aitem $f \setto LoadTimingModel()$
\Aitem $b^* \setto \operatorname*{argmin}_{b \in \mathcal{B}} f(a, b, c)$
\algoremark{\color{darkred}{fast}}
\Aitem \Return $b^*$
\end{algorithm}
\caption{Pseudo-code template for on-demand auto-tuning.
Empirical auto-tuning (above) is inevitably slow
because dynamically-generated code must be compiled and run on a number of actual-size inputs.
Predictive auto-tuning (below) can be orders of magnitude faster. We show that it can
also be accurate.
}
\label{fig:general_autotune}
\end{figure}

This work shows that auto-tuning can be accelerated by orders of
magnitude by using a regression model built offline as a surrogate for actual
computations on the real hardware.
The general form of an auto-tuning based library routine is illustrated in Figure~\ref{fig:general_autotune} (top).
We will use filterbank correlation (Section~\ref{sec:fbcorr}) as our running example, but the approach is both general and powerful.
An auto-tuning based routine must operate on three sets of variables:
\begin{description}
\item[$\mathcal{A}$:] task description (argument shapes, physical layout)
\item[$\mathcal{B}$:] implementation description (auto-tuning parameters)
\item[$\mathcal{C}$:] platform description (compute capability, micro-benchmarks)
\end{description}
The hypothetical auto-tuning routine described at the top of Figure~\ref{fig:general_autotune} 
might take many minutes or hours to perform the argmin at step 3 (during which time it computes the desired result many times!)
so it would be all but unusable as a normal library routine.
However, the form of the auto-tuning routine suggests the potential for enormous acceleration:
if only there were a fast (even approximate) surrogate for the costly $MeasureTime(\cdot)$ function,
then the argmin could be done in a fraction of a second and the routine could be used normally (Figure~\ref{fig:general_autotune}, bottom).

Our contribution of the present work is to show that boosted regression tree models,
a powerful machine learning technique for function approximation, can meet the
requirements of this application.
We call this approach {\em predictive} auto-tuning to contrast it with the standard measurement-based approach which we will call {\em empirical} auto-tuning.
XXX: We show that the kernel from the GCG chapter can be modeled on a variety of hardware and get good performance compared to empirical auto-tuning.

@@

\subsection{Filterbank Correlation}
\label{sec:fbcorr}

Filterbank correlation is a simple spatial image filtering operation that is
an important subroutine in many image processing applications. It has a relatively
high arithmetic intensity which makes it a natural fit for GPU platforms~\citep{pinto+cox:2011gcg}. XXX cite some computer vision papers.

Mathematically, we define filterbank correlation in terms of an
image $x$ and a filterbank $f$.
The image $x$ has $R$ rows, $C$ columns, and $D$ channels (e.g. color
channels) that we call its {\em depth}. We index $x$ like $\mathbf{x}[i,j,d]$
where $0 \leq i < R$, $0 \leq j < C$, and $0 \leq d < D$.
The filterbank $f$ has $F$ filters that are like little images: each has a
height $H$, a width $W$, and $D$ channels.
We will restrict ourselves to what are called {\em valid} correlations, in
which the image is larger in both rows and columns than the filters.
The result of filterbank correlation of $x$ with $f$ is an image-like array
$z$ with $R-H+1$ rows, $C-W+1$ columns, and depth $F$, whose elements are
defined according to Equation~\ref{eq:z}:

@@

In terms of floating point operations, a filterbank correlation requires the
inner sums to be computed for each output pixel, yielding the quantity in
Eq.~\ref{eq:flops}:
\begin{equation}
\mathrm{FLOPS} = 2  K  H  W  D  (R - H + 1)( C- W + 1)
\label{eq:flops}
\end{equation}
The multiplicative factor of 2 arises because we must first multiply an element of
$x$ with an element of $f$ and then add the result to an element of $z$.

The memory transfer requirements of filterbank correlation are more difficult to quantify.
Assuming three kinds of non-register memory -- constant, shared, and global --
and assuming optimistically that the entire filterbank fits into the GPU's constant memory,
then we can establish a lower bound (Eq.~\ref{eq:bytes}) on the amount of
memory that must be moved in order to store the computed result to global
memory starting from arguments in global memory:
\begin{align}
\mathrm{Bytes} = 4&RCD \nonumber \\
& + 4KHWD \nonumber \\
& + 4(R-H+1)(C-W+1)K.
\label{eq:bytes}
\end{align}
In short, we must read the filterbank and image once, and store the result.

The arithmetic intensity of filterbank correlation, assuming our lower bound on memory transfers
is therefore approximately
\begin{align}
\mathrm{intensity} & \approx \frac {KDHW} {2(D+K)},
\end{align}
for images that are large relative to filters.
Each $K$ output writes corresponds to approximately $D$ input reads
and $K$ inner products between $DHW$ elements.

The high potential for arithmetic intensity makes the GPU an ideal platform for computing filterbank correlations,
and and filterbank correlation is used extensively in image and video processing,
where it is often a computational bottleneck.
One might expect then, that it would be easy to implement a library providing
this operation -- this hypothetical library could have quite a simple interface
following the spirit of FFTW or a BLAS routine:
\begin{verbatim}
    gpu_fbcorr(
        img_shape, filters_shape,
        img, filters, output)
\end{verbatim}
However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
related stencil operations XXX, it is challenging to provide an implementation
or even an implementation strategy that provides satisfactory performance
across the range of inputs (shapes, physical layouts) that occur in
typical usage.

@@

\subsection{GPU Implementation}

@@

\section{Experiments}

Recall from the introduction (Eq.~\ref{eq:condopt}) that auto-tuning can be seen
as a conditional optimization problem. We wish to find an implementation ($\mathcal{B}$) that minimizes runtime or some other scalar-valued cost function
for given arguments ($\mathcal{A}$) on a particular platform ($\mathcal{C}$).
In order to perform predictive auto-tuning it is necessary to characterize these
three types of variables with {\em features}.
In Section~\ref{sec:modeling} we see how these features are used as inputs to
the regression model in predictive auto-tuning.

\subsection{Argument Description $\mathcal{A}$}

We describe the arguments to a filterbank correlation with the 6-tuple
$(R, C, D, F, H, W)$.
We randomly sampled problem specifications from the following space:
\begin{align*}
R = C & \in \{ 256, 512, 1024, 2048, 4096 \} \\
H = W & \in \{ 3, 5, 7, 9, 11 \} \\
D &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \} \\
F &  \in \{1, 4, 8, 16, 32, 64, 128, 256 \}
\end{align*}
The product space
includes 1600 elements, but we restricted our experiments to problems that
represented between 1 and 50 gigaflops. Our problem space included 602
configurations.

A library implementation that must support all image and
filter sizes must handle on the order of $2^{50}$ forseable
problem configurations (not even including variations due to strided memory
layouts).

@@

One might expect then, that it would be easy to implement a library providing
this operation -- this hypothetical library could have quite a simple interface
following the spirit of FFTW or a BLAS routine:
\begin{verbatim}
    gpu_fbcorr(
        img_shape, filters_shape,
        img, filters, output)
\end{verbatim}
However, as shown in \citet{pinto+cox:2011gcg} and numerous articles on
related stencil operations XXX, it is challenging to provide an implementation
or even an implementation strategy that provides satisfactory performance
across the range of inputs (shapes, physical layouts) that occur even in
typical usage.
